---
title: "Introduction to syncomR"
author: "Sudarshan A. Shetty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to syncomR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The `syncomR` is an R package consisting of functions for use with time-series microbial census data. The package is mainly accompanying the research article "Functional analysis of a synthetic gut microbiome reveals niche segregation as a driving force for assembly, stability, co-existence and activity" Shetty SA., et al 2020. Therefore, the functions have been tested using a synthetic minimal microbiome that was assembled in three bioreactors. Published R codes were wrapped for convenient handling of the temporal census data of synthetic minimal microbiome. Therefore, this package is considered as a research compendium for organizing and sharing codes and files pertaining to the said research article.    

## Installing syncomR   
```{r, eval=FALSE}
install.packages("devtools")

devtools::install_github("microsud/syncomR")
```

## Load libraries

```{r warning=FALSE, message=FALSE}
library(syncomR)
library(readr)
library(tidyr)
library(magrittr)
#library(edgeR)
library(RColorBrewer)
library(DT)
library(knitr)
library(patchwork)

fasting_cols <- c("#C71585", "#008080")

fer_cols <- c(Bioreactor_5= "#b2182b", Bioreactor_6="#2166ac", Bioreactor_8 = "#35978f")
```


# Convert to taxa-time table    

```{r taxa-time, message=FALSE, warning=FALSE}
ps1.b5 <- subset_samples(SyncomFiltData, StudyIdentifier == "Bioreactor A")
otu.tb <- taxa_time_table(ps1.b5, 
                          normalize = TRUE, 
                          time.col = "Time_hr", 
                          remove.zero = TRUE)
knitr::kable(otu.tb)
```

## Coverage    
Calculate how many taxa account for a certain percentage of total abundance in a community. Here, we set 95% threshold.  

```{r coverage, fig.height=4, fig.width=8, fig.align='center'}

data(SyncomFiltData)
fasting_cols <- c("#b3de69", "#fb8072", "#80b1d3")
pl <- taxa_coverage(SyncomFiltData,
                    coverage = 0.95,
                    time.column = "Time_hr_num",
                    shape.variable = "Acetate_Feed",
                    color.variable = "Fasting",
                    color.pal = fasting_cols,
                    y.breaks = seq(0, 16, 1),
                    y.limits = c(3, 13)
)
pl <- pl + facet_wrap(~StudyIdentifier) + theme_syncom() 
pl

```


## Community composition   

Visualizing overall community composition over time in different bioreactors.  

** Area plot**  
```{r Community-composition-area, fig.align='center', fig.height=6, fig.width=8}
data(SyncomFiltData)
SyncomFiltData.rel <- microbiome::transform(SyncomFiltData, "compositional")
p <- plot_syncom_composition(SyncomFiltData.rel,
                             type = "area",
                             time.col = "Time_hr_num",
                             taxa.level = "OTU",
                             sp.fill.pal = syncom_colors("BacterialSpecies"), 
                             facet.var = "StudyIdentifier")
print(p + theme_syncom())   
```

** Line plot**  

Here, we will use a combination of `add_time_lag` function and line plot. We want to see only those samples for Bioreactor A that are from fasting, i.e. samples taken after 12h or more gaps.  

```{r Community-composition-line, fig.align='center', fig.height=6, fig.width=16}
data(SyncomFiltData)
SyncomFiltData.rel <- microbiome::transform(SyncomFiltData, "compositional")
bioa_ps <- subset_samples(SyncomFiltData.rel, StudyIdentifier=="Bioreactor A")
fasting_ps <- add_time_lag(bioa_ps)
fasting_ps <- subset_samples(fasting_ps, time_lag >= 12)

p <- plot_syncom_composition(fasting_ps,
                             type = "bar",
                             time.col = "Time_hr_num",
                             taxa.level = "OTU",
                             sp.fill.pal = syncom_colors("BacterialSpecies"), 
                             #facet.var = "StudyIdentifier",
                             numeric.x.axis = F) 
print(p + theme_syncom())
```

## Temporal trajectory of individual taxa  
Now, we will look at individual taxa relative abundances over time.  

```{r Temp-tracjectory, fig.height = 3, fig.width = 6, fig.align='center'}

ps1.sycom.rel <- microbiome::transform(SyncomFiltData, "compositional")

fer_cols <- c(`Bioreactor A`= "#b2182b", `Bioreactor B`="#2166ac", `Bioreactor C` = "#35978f")

tax.trac <- plot_trajectory(ps1.sycom.rel,
                            time.col = "Time_hr",
                            taxa.level = "Species",
                            type = "single",
                            taxa = "Akkermansia_muciniphila",
                            group.variable = "StudyIdentifier",
                            color.pal = fer_cols,
                            transform.counts = "compositional")
tax.trac <- tax.trac + geom_smooth() + theme_syncom()
## a line to specific timepoint of interest
tax.trac + geom_vline(xintercept = 152, size=3, alpha=0.6, color="#e41a1c") 

```


## Temporal turnover of communities    

Let us now check for temporal patterns in community.  
```{r Temporal-turnover, fig.align='center'}
library(patchwork)
# Succession

data(SyncomFiltData)
ps.b5 <- subset_samples(SyncomFiltData, StudyIdentifier== "Bioreactor A")
p.start <- temporal_turnover(ps.b5, tree =NULL,
                             time.col = 'Time_hr_num', 
                             method = "canberra", 
                             compositional = TRUE,
                             compared.to = "start") + theme_syncom()
#p.start 

p.subseq <- temporal_turnover(ps.b5, tree =NULL,
                              time.col = 'Time_hr_num', 
                              method = "canberra", 
                              compositional = TRUE,
                              compared.to = "subsequent")+ theme_syncom() 
#p.subseq  

p.final <- temporal_turnover(ps.b5, tree =NULL,
                             time.col = 'Time_hr_num', 
                             method = "canberra", 
                             compositional = TRUE,
                             compared.to = "final") + theme_syncom() 
#p.final 

p.start / p.subseq / p.final

```

The temporal turnover data can be extracted as follows.  
```{r eval=FALSE}
# extract data 
dat <- p$data
kable(head(dat))
```


# Temporal changes in eveness of community  
Currently, `syncomR` only inequality from `microbiome` package is supported.   

```{r Temporal-eveness, fig.align='center', fig.width=4, fig.height=6}
data(SyncomFiltData)
p <- temporal_diversity(SyncomFiltData, 
                        time.col = 'Time_hr_num', 
                        div.measure = "gini")
p <- p + facet_wrap(~StudyIdentifier, ncol=1) + geom_smooth(fill = "#a6bddb", color = "#d94801") 
p + theme_syncom()

```

## Comunity-level stability properties  
Calculate community stability properties as decribed by Florian Centler and Zishu Liu, UFZ Leipzig, Germany. [Ecological Stability Properties. Microbial Communities Assessed by Flow Cytometry" by Liu et al., 2018 ](http://msphere.asm.org/content/3/1/e00564-17). Please cite this work when using this function. A value of -1 will let the script select these time points automatically, assuming that the file contains a single disturbance experiment (i.e. start and end of the experiment refer to the first and last entry, respectively). Otherwise, specify a time value which is actually present in the data, i.e. a value which appears in the first column of the input file. 
For more information check `?stability_properties`.  
```{r stability-properties}
data(SyncomFiltData)
ps1.b5 <- subset_samples(SyncomFiltData, StudyIdentifier== "Bioreactor A")
ps1.sub <- subset_samples(ps1.b5, Time_hr_num >= 12)
dat.stab <- stability_properties(ps1.sub, time.col = "Time_hr")# B. hydrogenotrophic introduced
knitr::kable(head(dat.stab))

```

## Community trajectory (2D ordination)  
Input is from the `stability_properties` function.  
 
```{r nmds-stability, fig.align='center', fig.width=6, fig.height=4}

p_ba <- plot_nmds_stability(dat.stab) 
# the nmds message is printed on purpose for user to check if Solution is reached
p_ba$plot + theme_syncom() + xlab("NMDS1") + ylab("NMDS2")

```

## Plot stability properties   

Distance overtime compared to refrence state.

```{r fig.align='center', fig.width=8, fig.height=4}
data(SyncomFiltData)
ps1.b5 <- subset_samples(SyncomFiltData, StudyIdentifier== "Bioreactor B")
dat.stab <- stability_properties(ps1.b5, time.col = "Time_hr",
                                 experimentStart = 24, # first spiked sample after initiation of continous feed operation
                                 tref=168)
plot_resistance(dat.stab) + geom_rect(aes(group = NULL,   col = NULL, ymax = -0.1, ymin = -0.2, xmin=120, xmax=470),
      fill = "#FDFEFE", 
      col = "#FDFEFE"
    ) + ylim(c(-0.2,0.6)) + xlim(c(120,470)) + coord_polar() + theme_bw() 

```

```{r fig.width=8, fig.height=6}

euc <- plot_resilience(dat.stab, method = "euclidean") + geom_path(alpha = 0.9, arrow = arrow(angle = 15, length = unit(0.1, "inches"),
                  ends = "last", type = "closed"), color="steelblue")
canb <- plot_resilience(dat.stab, method = "canberra") + geom_path(alpha = 0.9, arrow = arrow(angle = 15, length = unit(0.1, "inches"),
                  ends = "last", type = "closed"), color="brown")#+ ylim(c(0,0.6)) + coord_polar()

euc / canb
```

resistance resembles the ability of a system to remain unchanged despite a perturbation 
resilience describes the ability of a system to recover to a reference state after a perturbation

Online resilience overtime.  
```{r stability-properties-2, fig.align='center', fig.width=8, fig.height=3}

plot_resilience_online(dat.stab, col.low = "#fa9fb5", col.high = "#49006a") + geom_path(alpha = 0.9, arrow = arrow(angle = 15, length = unit(0.1, "inches"),
                  ends = "last", type = "closed"), color="red")+ 
  coord_polar(theta = "x") + theme_bw()  + ylim(c(-0.2,0.4))+ xlim(c(24, 460)) 
#+  ADD ARROW SHOWING DIRECTION FROM ZERO TO 460h
#+ geom_curve(aes(x = 200, y = 0.0, xend = 400, yend = 0.4), curvature = 4)  
```

## Gut metabolic modules  

```{r GMM-plot, fig.align='center', fig.width=6, fig.height=6}
data(SyncomGMM)
focal.species <- c("Bacteroides_ovatus", "Bacteroides_xylanisolvens")
p<- plot_module_abundances(SyncomGMM,
                           tax.variable = focal.species,
                           mm.variable = c("propionate production III", "lactose degradation","xylose degradation", "galactose degradation"),
                           color.pal = syncom_colors("BacterialSpecies"),
                           nrow=3,
                           ncol=1)
p + theme_syncom() + theme(legend.position="bottom")
``` 


```{r}
sessionInfo()
```



